---
title: "Multiple Regression"
output:
  github_document:
    html_preview: true
    df_print: paged
---

## Study

Last year, Brielle's company released a single-serving drink that facilitates weight loss. To measure the effectiveness of the product, the company invites website visitors at random to give a number of their total weight loss. From a pre-release controlled pilot study, Brielle found evidence that gender and height may influence weight loss, so the survey also asked for those pieces of information. 

Brielle is also interested in knowing how these self-report weights match up to the differences in weights during the company's pilot experiment. She believes that the difference in these study results can provide an idea of potential descrepancies in generalizaing the controlled experiment (holding other factors still such as diet, exercise, daily activities, etc.) to a real-world setting. 

## Data simulation

```{r}

```


## Why this model?
We want to use multiple regression here for a few reasons:

* 

```{r}
set.seed(150)

##### Simulate data ---------------
lower.sales <- runif(n = 15, min = 0, max = 2)
upper.sales <- runif(n = 15, min = 2, max = 4)

df <- data.frame(
  Quantity = append(upper.sales, lower.sales),
  Price = seq(2, 3.45, .05)
)
```

## Visualization
First, we want to look at the data graphically.

```{r}
library(ggplot2)
theme_set(theme_bw())

p <- ggplot() +
  geom_point(aes(x = Price, y = Quantity), df) +
  labs(
    x = "Unit Price \n (U.S. Dollars)",
    y = "Quantity Sold \n (hundreds)"
  ) 
p
```

Once we model the data, our intercept will represent the quantity sold when the price is 0. Tom is not necessarily interested in how his hypothesis or model applies to the general population. Therefore, we want to center the Price variable so the intercept represents the change from our initial price.

```{r}
df$Price.c <- df$Price - 2.75 
```


## Model 

There appears to be a negative relationship between quantity sold and unit price. Let's see how a simple linear model would look. Note how our centering changes the meaning of alterations in price, but does not affect the spread of our values.

```{r}
### Make model and predictions -----
mod1 <- lm(Quantity ~ Price.c, data = df)
mod1.summ <- summary(mod1)
df$preds <- predict(mod1, df)

### Plot regression line -----
ggplot() +
  geom_line(aes(x = Price.c, y = preds), color = 'red', df) +
  geom_point(aes(x = Price.c, y = Quantity), df) +
    labs(
    x = "Change from Starting Price \n (U.S. Dollars)",
    y = "Quantity Sold \n (hundreds)"
  ) 
```

Using the centered predictor variable, our model will now represent how change in price from product A's initial sale price affects quantity sold. Let's model it and go through the most important parts of the results. 

```{r}
options(scipen=1, digits=3) # number formatting

### Model
summary(mod1) 

### Confidence interval
cints <- confint(mod1, c('(Intercept)', 'Price.c'), level=0.95)
cints
```

## Interpretation

#### Intercept
* When the price is $2.75, the predicted quantity sold is `r abs(mod1.summ$coefficients[[1]])` (hundred) units
* If we were to replicate this experiment under identical conditions, there is a 95% chance that the change in quantity sold for increasing pricing by an additional dollar would be between `r cints['(Intercept)',][1]` and `r cints['(Intercept)',][2]`

#### Price
* Price is a statistically significant predictor of quantity sold (p < .001)
* For every dollar that the price increases from our $2.75 starting point, the quantity sold is expected to drop by `r abs(mod1.summ$coefficients[[1]])` (hundred) units
* If we replicated this experiment, there is a 95% chance that the change in quantity sold for increasing pricing by an additional dollar would be between `r cints['Price.c',][1]` and `r cints['Price.c',][2]`

#### Model
* Our full model explains about 48% of the variance in the quantity sold
