---
title: "Simple Regression"
output:
  github_document:
    html_preview: true
    df_print: paged
---

## Study
Tom wants to explore and model how adjusting the pricing of product A affects the quantity sold. The initial price of a single unit is $2.75.

```{r}
set.seed(150)

##### Simulate data ---------------
lower.sales <- runif(n = 15, min = 0, max = 2)
upper.sales <- runif(n = 15, min = 2, max = 4)

df <- data.frame(
  Quantity = append(upper.sales, lower.sales),
  Price = seq(2, 3.45, .05)
)
```

## Visualization
First, we want to look at the data graphically.

```{r}
library(ggplot2)
theme_set(theme_bw())

p <- ggplot() +
  geom_point(aes(x = Price, y = Quantity), df) +
  labs(
    x = "Unit Price \n (U.S. Dollars)",
    y = "Quantity Sold \n (hundreds)"
  ) 
p
```

Once we model the data, our intercept will represent the quantity sold when the price is 0. This could be problematic in interpretation, because the price never hits 0. In fact, it stays between $2 and $3.50. Therefore, we want to change the Price values to something more meaningful: to represent the change from our initial price.

```{r}
df$Price.c <- df$Price - 2.75 
```


## Model 

There appears to be a negative relationship between quantity sold and unit price. Let's see how simple linear model would look. Note how our centering changes the meaning of alterations in price, but does not affect the spread of our values.

```{r}
### Make model and predictions -----
mod1 <- lm(Quantity ~ Price.c, data = df)
mod1.summ <- summary(mod1)
df$preds <- predict(mod1, df)

### Plot regression line -----
ggplot() +
  geom_line(aes(x = Price.c, y = preds), color = 'red', df) +
  geom_point(aes(x = Price.c, y = Quantity), df) +
    labs(
    x = "Change from Starting Price \n (U.S. Dollars)",
    y = "Quantity Sold \n (hundreds)"
  ) 
```

Using the centered predictor variable, our model will now represent how change in price from product A's initial sale price affects quantity sold. Let's model it and go through the most important parts of the results. 

```{r}
options(scipen=1, digits=3) # number formatting

### Model
summary(mod1) 

### Confidence interval
cints <- confint(mod1, 'Price.c', level=0.95)
cints
```

Our model would be represented as:

<i> Quantity = `r mod1.summ$coefficients[[1]]` + (`r mod1.summ$coefficients[[2]]`) + Error </i>

## Interpretation
* Price is a statistically significant predictor of quantity sold
* When the price is $2.75, the predicted quantity sold is `r abs(mod1.summ$coefficients[[2]])` (hundred) units
* For every dollar that the price increases from our $2.75 starting point, the quantity sold is expected to drop by `r abs(mod1.summ$coefficients[[1]])` (hundred) units
* If we were to accurately replicate this experiment, there would be a 95% chance that the change in quantity sold for increasing pricing by an additional dollar would be between `r cints[1]` and `r cints[2]`
* Our model explains about 48% of the variance in quantity sold
