---
title: "One-way ANOVA"
output:
  github_document:
    html_preview: true
    df_print: paged
---

## Study
Tim is a pharmacology researcher designing a study to test the effectiveness of drugs A and B on treating social anxiety in rat subjects with the disorder. He randomly assigns subjects to 3 groups: Placebo (control), Drug A, and Drug B. That is; 14 subjects receive a sugar pill, 14 receive Drug A, and 14 receive Drug B.

Each subject is given a 2-minute socialization period with a juvenile subject. The percentage of time each subject spends exhibiting "social behaviors" with the juvenile is recorded. 

Specifically, Tim wants to know if:

1. Giving subjects a drug treatment reduces social anxiety, especially the extent to which it does or does not. If the drug is to be considered "effective", time socializing in the drug groups will be higher than the placebo group.
2. One drug treatment is more effective than the other

## Data simulation

We simulate data from the random uniform distribution.

```{r}
set.seed(1500)
options(scipen=1, digits=3) # number formatting

##### Simulate data ---------------
Placebo <- runif(n = 14, min = 20, max = 45)
drugA <- runif(n = 14, min = 57, max = 76)
drugB <- runif(n = 14, min = 52, max = 78)

df <- data.frame(
  Perc.social = append(append(Placebo, drugA), drugB),
  Group = append(append(rep('Placebo', 14), 
                 rep('Drug A', 14)),
                 rep('Drug B', 14)))

##### Format ---------------
df$Group <- factor(df$Group, levels = c('Placebo', 'Drug A', 'Drug B')) # reorder
```

First, let's look at the data. We can look at both the distribution of data points and quartile information by using a boxplot. 

```{r}
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
theme_set(theme_bw())

##### Plot ---------------
p <- ggplot() +
  geom_point(aes(x = Group, y = Perc.social), df) +
  geom_boxplot(aes(x = Group,
                   y = Perc.social), alpha = 0.5, df) +
  labs(x = "Group", y = "Percent of time Socializing")

p
```

## Model 

Before we construct the model, we design contrasts to represent group differences. These contrasts test Tim's hypotheses 1 and 2, respectively. Is there an effect of drug, and does this effect differ between drugs?

```{r}
##### Contrasts ---------------
con.1 <- c(-1, 0.5, 0.5) # Placebo vs. drug
con.2 <- c(0, -0.5, 0.5) # Drug A vs. Drug B

con.mat <- rbind(con.1, con.2)

rownames(con.mat) <- c('drug', 'A vs B')
colnames(con.mat) <- c('Placebo', 'Drug A', 'Drug B')

con.mat

```

We give each contrast their own column in the main dataframe by multiplying values of each contrast by their respective outcomes.

```{r}
source('mcSummary.R')
suppressMessages(library(car))
suppressMessages(library(psych))
suppressMessages(library(lmSupport))

##### Apply contrasts ---------------
df$drug.con <- (df$Group == 'Placebo')*(-1) + (df$Group == 'Drug A')*(0.5) + (df$Group == 'Drug B')*(0.5)

df$AB.con <- (df$Group == 'Placebo')*(0) + (df$Group == 'Drug A')*(-0.5) + (df$Group == 'Drug B')*(0.5)

##### Model data ---------------
mod1 <- lm(Perc.social ~ drug.con + AB.con, data = df)
mod1.summ <- mcSummary(mod1)

mod1.summ
```


## Interpretation

#### Intercept
* On average, subjects spent `r mod1$coefficients[[1]]` percent of the 2-minute test period exhibiting social behavior
* If we replicated this study, there is a 95% chance that the mean percent of social time of all subjects would be between `r mod1.summ$coefficients[,'CI_2.5'][[1]]` and `r mod1.summ$coefficients[,'CI_97.5'][[1]]`

#### Contrast 1
* Subjects that received a drug treatment spent an average of `r mod1$coefficients[[2]]` percent more time socializing than those who did not receive a drug
  * This difference is statistically significant
* If we were to replicate this study, there is a 95% chance that the difference in means between the Placebo group and drug groups would between `r mod1.summ$coefficients[,'CI_2.5'][[2]]` and `r mod1.summ$coefficients[,'CI_97.5'][[2]]`
* Drug intake explains 94.3% of variance in time spent socializing

#### Contrast 2
* Of subjects that received a drug, ones that received Drug B spent an average of `r mod1$coefficients[[3]]` percent more time socializing than those who received a drug 
  * This difference is not statistically significant
* There is a 95% chance that if we replicated our experiment, the difference in means between subjects that received Drug A and subjects that received Drug B is between `r mod1.summ$coefficients[,'CI_2.5'][[3]]` and `r mod1.summ$coefficients[,'CI_97.5'][[3]]`
* The difference in drugs explains 12.6% of the variance in time spent socializing

## Visualization

To visualize our summary, we can convert the coefficients from our summary object into a data frame. We also add label variables for our plot. 

```{r}
### Extract coefficients
co <- as.data.frame(mod1.summ$coefficients) # extract coefficients

##### Create labels --------------
co$Comparison <- rownames(co)
comp.labs <- c('(Intercept)', 'Drug - No Drug', 'Drug B - Drug A')
co$Comp.labs <- factor(comp.labs, levels = comp.labs)

co <- co %>% filter(Comparison == 'drug.con' | Comparison == 'AB.con')

```

```{r}
##### Plot --------------
results.plot <- ggplot(co) +
  geom_point(aes(x = Comp.labs, y = Est)) +
  geom_errorbar(aes(x = Comp.labs,
                    ymin = CI_2.5,
                    ymax = CI_97.5)) +
  scale_x_discrete(labels = co$Comp.labs,
                   breaks = co$Comp.labs,
                   limits = rev(co$Comp.labs)) +
  coord_flip()

##### Add labels and theme --------------
results.plot <- results.plot + 
  labs(x = '', y = 'Mean Difference',
    title = 'Contrasts of Mean Differences',
    subtitle = 'with confidence intervals') +
  theme(
    axis.title.x = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.minor.x = element_line(color = 'gray96'),
    panel.grid.major.x = element_line(color = 'gray96'),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.text.x = element_text(face = 'bold'),
    axis.text.y = element_text(face = 'bold')
  )

results.plot
```
