---
title: "t-test"
output:
  github_document:
    html_preview: true
    df_print: paged
---

## Study
Sarah, an HR rep, is looking to improve employee satisfaction in Clinic A. She hypothesizes that these employee satisfaction ratings can be raised by adding a 15-minute socialization period in the afternoon. To perform this, Sarah randomly selects half of the staff to receive the socialization period (social group) and half to carry on business as usual (control group).

```{r}
set.seed(150)
options(scipen=1, digits=3) # number formatting

##### Simulate data ---------------
cont.grp <- runif(n = 20, min = 4, max = 10)
social.grp <- runif(n = 20, min = 6, max = 10)

df <- data.frame(
  Rating = append(cont.grp, social.grp),
  Group = append(rep('control', 20), rep('social', 20))
)
```

First, let's look at the data. We can look at both the summary statistics and the values by adding individual points to a bar graph of the means.

```{r}
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
theme_set(theme_bw())

summ.df <- df %>% 
  group_by(Group) %>% 
  summarize(Rating = mean(Rating))

# percent change
perc.diff <- (summ.df$Rating[summ.df$Group == 'social'] - summ.df$Rating[summ.df$Group == 'control']) / abs(summ.df$Rating[summ.df$Group == 'control'])
perc.diff <- perc.diff*100

p <- ggplot() +
  geom_bar(aes(x = Group, y = Rating), summ.df, stat = 'identity') +
  geom_point(aes(x = Group, y = Rating), df) +
  labs(
    x = "Group",
    y = "Employee Satisfaction Rating"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
p
```

The social group seems to rate satisfaction higher than the control group. However, the difference is so small that it looks like there's a good chance it could also be the result of coincidence. This is where statistical modeling helps by giving us an idea of whether a socialization period was a significantly effective manipulation in increasing average employee satisfaction ratings and how big of an effect it had and may have if our experiment is replicated. 

## Model 

To model this, we can simply use dummy coding in a regression context.

```{r}
### Code variables
df$Group.code <- ifelse(df$Group == 'control', 0, 1)
```

By dummy coding our groups, we can interpret our results in terms of how the presence of a socialization period is related to employee satisfaction ratings.

```{r}
### Make model and predictions -----
mod1 <- lm(Rating ~ Group.code, data = df)
mod1.summ <- summary(mod1)

mod1.summ

df$preds <- predict(mod1, df)
```

We almost achieve statistical significance with p > .05. In context, however, an R-squared value of `r mod1.summ$r.squared` is somewhat interesting, but the actual decision to add the break period depends on the opportunity cost for taking away 15 minutes from working.

```{r}
### Model
summary(mod1) 

### Confidence interval
cints <- confint(mod1, 'Group.code', level=0.95)
cints

```

Our model would be represented as:

<i> Rating = `r mod1.summ$coefficients[[1]]` + (`r mod1.summ$coefficients[[2]]`)Group + Error </i>

## Interpretation
* A socialization period (Group) resulted in an almost-statistically significant difference in employee satisfaction ratings (p = `r mod1.summ$coefficients[[8]]`)
* In the absence of a socialization period, ratings are, on average, `r abs(mod1.summ$coefficients[[1]])`
* When a socialization period is present, employee satisfaction ratings are expected to be `r abs(mod1.summ$coefficients[[2]])` points higher than those who do not receive one; a difference of about `r perc.diff` 
* If we were to accurately replicate this experiment, there is a 95% chance that the change in ratings from the addition of a socialization period is between `r cints[1]` and `r cints[2]`
* Our model explains about `r mod1.summ$r.squared` of the variance in employee satistfaction ratings
