---
title: "Multiple Regression"
output:
  github_document:
    html_preview: true
    df_print: paged
---

## Study
Earl, a restaurant owner, wants to estimate the number of days customers will go between visits.

```{r}
set.seed(150)

##### Simulate data ---------------
lower.sales <- runif(n = 15, min = 0, max = 2)
upper.sales <- runif(n = 15, min = 2, max = 4)

df <- data.frame(
  Quantity = append(upper.sales, lower.sales),
  Price = seq(2, 3.45, .05)
)
```

## Visualization
First, we want to look at the data graphically.

```{r}
library(ggplot2)
theme_set(theme_bw())

p <- ggplot() +
  geom_point(aes(x = Price, y = Quantity), df) +
  labs(
    x = "Unit Price \n (U.S. Dollars)",
    y = "Quantity Sold \n (hundreds)"
  ) 
p
```

Once we model the data, our intercept will represent the quantity sold when the price is 0. Tom is not necessarily interested in how his hypothesis or model applies to the general population. Therefore, we want to center the Price variable so the intercept represents the change from our initial price.

```{r}
df$Price.c <- df$Price - 2.75 
```


## Model 

There appears to be a negative relationship between quantity sold and unit price. Let's see how a simple linear model would look. Note how our centering changes the meaning of alterations in price, but does not affect the spread of our values.

```{r}
mod.allison = 
  coxph(Surv(week, arrest) ~ fin + age + race + wexp + mar + paro + prio, 
        data=Rossi)
summary(mod.allison)
```

Using the centered predictor variable, our model will now represent how change in price from product A's initial sale price affects quantity sold. Let's model it and go through the most important parts of the results. 

```{r}
options(scipen=1, digits=3) # number formatting

### Model
summary(mod1) 

### Confidence interval
cints <- confint(mod1, c('(Intercept)', 'Price.c'), level=0.95)
cints
```

## Interpretation

#### Intercept
* When the price is $2.75, the predicted quantity sold is `r abs(mod1.summ$coefficients[[1]])` (hundred) units
* If we were to replicate this experiment under identical conditions, there is a 95% chance that the change in quantity sold for increasing pricing by an additional dollar would be between `r cints['(Intercept)',][1]` and `r cints['(Intercept)',][2]`

#### Price
* Price is a statistically significant predictor of quantity sold (p < .001)
* For every dollar that the price increases from our $2.75 starting point, the quantity sold is expected to drop by `r abs(mod1.summ$coefficients[[1]])` (hundred) units
* If we replicated this experiment, there is a 95% chance that the change in quantity sold for increasing pricing by an additional dollar would be between `r cints['Price.c',][1]` and `r cints['Price.c',][2]`

#### Model
* Our full model explains about 48% of the variance in the quantity sold
